---
output:
  md_document:
    variant: gfm
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
set.seed(0)
```

# EDMAinR - Euclidean Distance Matrix Analysis in R

[![CRAN version](http://www.r-pkg.org/badges/version/EDMAinR)](http://cran.rstudio.com/web/packages/EDMAinR/index.html)
[![CRAN download stats](http://cranlogs.r-pkg.org/badges/grand-total/EDMAinR)](https://www.rdocumentation.org/packages/EDMAinR/)
[![License: GPL v2](https://img.shields.io/badge/License-GPL%20v2-blue.svg)](https://www.gnu.org/licenses/old-licenses/gpl-2.0.en.html)

[![Linux build Status](https://travis-ci.org/psolymos/EDMAinR.svg?branch=master)](https://travis-ci.org/psolymos/EDMAinR)
[![Windows build status](https://ci.appveyor.com/api/projects/status/5y5fwgv90f8i84ck?svg=true)](https://ci.appveyor.com/project/psolymos/EDMAinR)
[![codecov](https://codecov.io/gh/psolymos/EDMAinR/branch/master/graph/badge.svg)](https://codecov.io/gh/psolymos/EDMAinR)

> A coordinate‐free approach for comparing biological shapes using landmark data

## Install

```{r eval=FALSE}
remotes::install_github("psolymos/EDMAinR")
```

See what is new in the [NEWS](NEWS.md) file.

## License

[GPL-2](https://www.gnu.org/licenses/old-licenses/gpl-2.0.html)

## Contributing

Feedback and contributions are welcome:

- submit feature request or report issues [here](https://github.com/psolymos/EDMAinR/issues),
- fork the project and submit pull request, see [CoC](CODE_OF_CONDUCT.md).

## Usage

```{r}
library(EDMAinR)

file1 <- system.file("extdata/crouzon/Crouzon_P0_Global_MUT.xyz",
    package="EDMAinR")
x1 <- read_xyz(file1)
x1

file2 <- system.file("extdata/crouzon/Crouzon_P0_Global_NON-MUT.xyz",
    package="EDMAinR")
x2 <- read_xyz(file2)
x2

B <- 9

fit <- edma_fit(x1, B=B)
fit
```

## References

Lele, S. R., 1991.
Some comments on coordinate-free and scale-invariant
methods in morphometrics.
American Journal of Physical Anthropology 85:407--417.
<doi:10.1002/ajpa.1330850405>

Lele, S. R., and Richtsmeier, J. T., 1991.
Euclidean distance matrix analysis: A coordinate-free approach for
comparing biological shapes using landmark data.
American Journal of Physical Anthropology 86(3):415--27.
<doi:10.1002/ajpa.1330860307>

Lele, S. R., and Richtsmeier, J. T., 1992.
On comparing biological shapes: detection of influential landmarks.
American Journal of Physical Anthropology 87:49--65.
<doi:10.1002/ajpa.1330870106>

Lele, S. R., and Richtsmeier, J. T., 1995.
Euclidean distance matrix analysis: confidence intervals for
form and growth differences.
American Journal of Physical Anthropology 98:73--86.
<doi:10.1002/ajpa.1330980107>

Hu, L., 2007. Euclidean Distance Matrix Analysis of Landmarks Data:
Estimation of Variance. Thesis, Master of Science in Statistics,
Department of Mathematical and Statistical Sciences,
University of Alberta, Edmonton, Alberta, Canada. Pp. 49.
