---
title: "Parametric estimation"
author: "Peter Solymos"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parametric estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup,include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(123)
library(EDMAinR)
```

## Introduction

This tutorial describes how to estimate SigmaK and assess if it is estimable.
Parametric estimation of SigmaK involves the following steps:

1. non-parametric estimation of mean form and SigmaKstar using the `edma_fit` function,
2. constructing of a pattern matrix based on biology of landmarks and respecting statistical constraints,
3. use the `SigmaK_fit` to estimate SigmaK using the non-parametric fit and the pattern matrix,
4. make sure the results are estimable by running `sensitivity` analysis.

## Simulated landmark data

To demonstrate the workflow, let us simulate some data.
Simulation is useful for checking the limits of the estimating procedure.
We know what the 'true' input are, because we are controlling those
through the simulation. We can then compare the outputs (our estimates),
to the known imputs, and this is how we evaluate the utility 
and limitations of our methods.

Let us begin with the following set of 6 landmarks in 2-dimensions:

```{r}
(M <- array(
  c(82, 21, 22, -50, -37, -37, 
  0,  17, -17, 0,  41,  -41), 
  dim=c(6, 2),
  dimnames=list(paste0("L", 1:6), c("X", "Y"))))
plot(M, pch=3, asp=1, axes=FALSE, ann=FALSE, type="n")
text(M, labels=rownames(M), col=4)
```

The simplest SigmaK variance-covariance matrix is $\sigma^2 I$, that is
constant variance ($\sigma^2$) in the diagonal of the $6\times6$ square
matrix and 0's in all the off-diagonal cells:

```{r}
sigma <- 2
SigmaK <- diag(sigma^2, 6, 6)
dimnames(SigmaK) <- list(rownames(M), rownames(M))
print_tb(SigmaK)
```

We can use the `edma_simulate_data` function to simulate `n`
specimen given form matrix `M` and covariance matrix `SigmaK`:

```{r}
n <- 1000
sim <- edma_simulate_data(n=n, M, SigmaK)

tmp <- plot_2d(sim, which=NULL, ask=NA,
  xlim=c(-80, 100), ylim=c(-60, 60), asp=1)
xy <- attr(tmp, "coordinates")
text(xy[,1]+15, xy[,2], labels=rownames(xy), cex=0.6, col=4)
```

We can see from the figure that the constant variance stranslates to
roughly circle shaped convex hulls around the landmarks
(the hull bounds the landmarks form the `n` specimens),
and the size of these patches is similar (equal variances).

## Nonparametruc estimation

We use the non-parametric EDMA algorithm to estimate
the form matrix and the SigmaKstar matrix:

```{r}
fit <- edma_fit(sim)
```

Because we used simulations, we can compare the known `M` matrix
with the estimated mean form:

```{r}
M                       # true
round(Meanform(fit), 2) # estimate
```

The 'true' `SigmaKstar` matrix is stored in our simulation object:

```{r}
tmp <- attr(sim, "simulation_settings")$SigmaKstar
round(tmp, 2)
round(SigmaKstar(fit), 2)
```

## Pattern matrix

The parametric estimation requires a pattern matrix
that describes the unknowns in the `SigmaK` matrix.
This `SigmaK` matrix is the variance covariance matrix
describing the variances (diagonal) associated with each landmark,
and the covariances (off-diagonal) among the landmarks.

The pattern matrix can be constructed in R, but when the number
of landmarks is high and variances are expected to differ among
the landmarks, it might be easier to construct the pattern matrix
in a spreadsheet editor (Excel). Here is how the file should be
constructed (the package contains example files, see
`?read_pattern`):

- the 1st row describes the landmarks starting from the 2nd column,
- the 1st column describes the landmarks in the same order as the column headers,
- leave the cell in the topleft corner empty,
- use unique names or numbers in the rest of the table to denote unique variables in the Sigma matrix,
- structural zeros (cells where the covariance is known or assumed to be 0) have to be blank,
- there cannot be blank cells in the diagonal of the matrix,
- values below and above the diagonal must be symmetric (value in row 2/column 3 must be the same as row 3/column 2), either unique names or numbers, or blank.

The pattern matrix for the equal variance case looks like this:

```{r}
p <- matrix(
  c("1", NA, NA, NA,  NA,  NA,
    NA, "1", NA, NA,  NA,  NA,
    NA, NA, "1", NA,  NA,  NA,
    NA,  NA,  NA, "1", NA, NA,
    NA,  NA,  NA, NA, "1", NA,
    NA,  NA,  NA, NA, NA, "1"),
  nrow=6, ncol=6, byrow=TRUE)
dimnames(p) <- list(rownames(M), rownames(M))
print_tb(p)
```

## Parametric estimation

The parametric estimation requires the fitted object from
the previous step of nonparamatric estimation
(that gave us the estimate of the mean form and SigmaKstar),
and the pattern matrix, so that the program knows what parameters
to estimate and where those are placed in the SigmaK matrix:

```{r}
(o <- SigmaK_fit(fit, p))
```

Now let us compare the 'true' SigmaK matrix with the estimated one:


```{r}
print_tb(SigmaK)              # true
print_tb(round(SigmaK(o), 2)) # estimate
```

## Sensitivity analysis

The parametric estimate is based on an optimization algorithm.
This algorithm requires starting values for the unknown parameters.
If a given parametric model's parameters are estimable,
we end up with the exact same estimates no matter what starting values
we use. This can be used to check the estimability of our parameters.
The `sensitivity` function repeats the parametric estimation
multiple times using different starting values:

```{r}
(s <- sensitivity(o))
```

The `par_1` column shows the parameter estimates for the variance,
the `value` column shows the value of the loss function evaluated
at `par_1`. Both these values are constant, which is a sign of estimability.

## Unequal variance case

In the unequal variance case we can have different variances for the
different landmarks. Here is the corresponding pattern matrix:

```{r}
p <- matrix(
  c("1", NA,  NA, NA,  NA,  NA,
    NA, "1", NA, NA,  NA,  NA,
    NA,  NA, "1", NA,  NA,  NA,
    NA,  NA,  NA, "2", NA,  NA,
    NA,  NA,  NA, NA, "2", NA,
    NA,  NA,  NA, NA,  NA, "2"),
  nrow=6, ncol=6, byrow=TRUE)
dimnames(p) <- list(rownames(M), rownames(M))
plot_tb(p)
```

The SigmaK matrix looks like this:

```{r}
parm <- c("1"=12, "2"=8)
SigmaK <- EDMAinR:::.vec2mat(parm, EDMAinR:::.mat2fac(p))
dimnames(SigmaK) <- dimnames(p)
print_tb(SigmaK)
```

The unequal variances translate to the following simulated data:

```{r}
sim <- edma_simulate_data(n=n, M, SigmaK)

tmp <- plot_2d(sim, which=NULL, ask=NA,
  xlim=c(-80, 100), ylim=c(-60, 60), asp=1)
xy <- attr(tmp, "coordinates")
text(xy[,1]+15, xy[,2], labels=rownames(xy), cex=0.6, col=4)
```

We can see the different sizes of the convex hulls corresponding to the
variances for each landmark.

For the estimation, we use `edma_fit` and `SigmaK_fit` as before:

```{r}
fit <- edma_fit(sim)
o <- SigmaK_fit(fit, p)
print_tb(SigmaK)              # true
print_tb(round(SigmaK(o), 2)) # estimate
round(s <- sensitivity(o), 3)
```

## Correlations

We can add correlations between the frontal (L1) and lateral (L2, L3)
landmarks, and similarly to the other 3 landmarks:

```{r}
p <- matrix(
  c("1", "2", "2", NA,  NA,  NA,
    "2", "1", NA,  NA,  NA,  NA,
    "2", NA,  "1", NA,  NA,  NA,
    NA,  NA,  NA,  "3", "4", "4",
    NA,  NA,  NA,  "4", "3", NA,
    NA,  NA,  NA,  "4", NA,  "3"),
  nrow=6, ncol=6, byrow=TRUE)
dimnames(p) <- list(rownames(M), rownames(M))
plot_tb(p)
```

The SigmaK matrix looks like this:

```{r}
parm <- c("1"=12, "2"=1, "3"=8, "4"=2)
SigmaK <- EDMAinR:::.vec2mat(parm, EDMAinR:::.mat2fac(p))
dimnames(SigmaK) <- dimnames(p)
print_tb(SigmaK)
```

The unequal variances translate to the following simulated data:

```{r}
sim <- edma_simulate_data(n=n, M, SigmaK)

tmp <- plot_2d(sim, which=NULL, ask=NA,
  xlim=c(-80, 100), ylim=c(-60, 60), asp=1)
xy <- attr(tmp, "coordinates")
text(xy[,1]+15, xy[,2], labels=rownames(xy), cex=0.6, col=4)
```

This setup seems estimable:

```{r}
fit <- edma_fit(sim)
o <- SigmaK_fit(fit, p)
print_tb(SigmaK)              # true
print_tb(round(SigmaK(o), 2)) # estimate
round(s <- sensitivity(o), 3)
```

## Non-estimable case: equi-correlation

The following parametrization is non identifiable:

```{r}
p <- matrix(
  c("1", "2", "2", "2", "2", "2",
    "2", "1", "2", "2", "2", "2",
    "2", "2", "1", "2", "2", "2",
    "2", "2", "2", "1", "2", "2",
    "2", "2", "2", "2", "1", "2",
    "2", "2", "2", "2", "2", "1"),
  nrow=6, ncol=6, byrow=TRUE)
dimnames(p) <- list(rownames(M), rownames(M))
plot_tb(p)
```

The diagonal elements are $\sigma^2$ (equal variance), the off diagonal
elements are $\sigma^2\rho$ where $\rho$ is the correlation:

```{r}
parm <- c("1"=10, "2"=2)
SigmaK <- EDMAinR:::.vec2mat(parm, EDMAinR:::.mat2fac(p))
dimnames(SigmaK) <- dimnames(p)
print_tb(SigmaK)
```

The correlation lead to more ellipsoid shapes for the simulated landmarks:

```{r}
sim <- edma_simulate_data(n=n, M, SigmaK)

tmp <- plot_2d(sim, which=NULL, ask=NA,
  xlim=c(-80, 100), ylim=c(-60, 60), asp=1)
xy <- attr(tmp, "coordinates")
text(xy[,1]+15, xy[,2], labels=rownames(xy), cex=0.6, col=4)
```

```{r}
fit <- edma_fit(sim)
o <- SigmaK_fit(fit, p)
print_tb(SigmaK)              # true
print_tb(round(SigmaK(o), 2)) # estimate
(s <- sensitivity(o))
```

The sensitivity analysis reveals that we get different estimates
when using different starting values. Although the estimates differ
greatly, the loss function values are identical.
Therefore, we cannot decide which estimate is the best,
these are equally bad, i.e. non-estimable.

## Non-estimable case: blocks

We can add correlations between the frontal (L1) and lateral (L2, L3)
landmarks, and similarly to the other 3 landmarks:

```{r}
p <- matrix(
  c("1", "2", "2", NA,  NA,  NA,
    "2", "1", "2",  NA,  NA,  NA,
    "2", "2", "1", NA,  NA,  NA,
    NA,  NA,  NA,  "3", "4", "4",
    NA,  NA,  NA,  "4", "3", "4",
    NA,  NA,  NA,  "4", "4","3"),
  nrow=6, ncol=6, byrow=TRUE)
dimnames(p) <- list(rownames(M), rownames(M))
plot_tb(p)
```

The SigmaK matrix looks like this:

```{r}
parm <- c("1"=12, "2"=1, "3"=8, "4"=2)
SigmaK <- EDMAinR:::.vec2mat(parm, EDMAinR:::.mat2fac(p))
dimnames(SigmaK) <- dimnames(p)
print_tb(SigmaK)
```

The unequal variances translate to the following simulated data:

```{r}
sim <- edma_simulate_data(n=n, M, SigmaK)

tmp <- plot_2d(sim, which=NULL, ask=NA,
  xlim=c(-80, 100), ylim=c(-60, 60), asp=1)
xy <- attr(tmp, "coordinates")
text(xy[,1]+15, xy[,2], labels=rownames(xy), cex=0.6, col=4)
```

This setup seems estimable:

```{r}
fit <- edma_fit(sim)
o <- SigmaK_fit(fit, p)
print_tb(SigmaK)              # true
print_tb(round(SigmaK(o), 2)) # estimate
round(s <- sensitivity(o), 3)
```


