---
title: "Parametric estimation"
author: "Peter Solymos"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parametric estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup,include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(123)
```

## Introduction

This tutorial describes how to estimate SigmaK and assess if it is estimable.
Parametric estimation of SigmaK involves the following steps:

1. non-parametric estimation of mean form and SigmaKstar using the `edma_fit` function,
2. constructing of a pattern matrix based on biology of landmarks and respecting statistical constraints,
3. use the `SigmaK_fit` to estimate SigmaK using the non-parametric fit and the pattern matrix.

```{r}
library(EDMAinR)
```

## Simulated landmark data

To demonstrate the workflow, let us simulate some data.
Simulation is useful for checking the limits of the estimating procedure.
We know what the 'true' input are, because we are controlling those
through the simulation. We can then compare the outputs (our estimates),
to the known inputs, and this is how we evaluate the utility 
and limitations of our methods.

Let us begin with the following set of 6 landmarks in 2-dimensions:

```{r}
(M <- array(
  c(82, 21, 22, -50, -37, -37, 
  0,  17, -17, 0,  41,  -41), 
  dim=c(6, 2),
  dimnames=list(paste0("L", 1:6), c("X", "Y"))))
plot(M, pch=3, asp=1, axes=FALSE, ann=FALSE, type="n")
text(M, labels=rownames(M), col=4)
```

The simplest SigmaK variance-covariance matrix is $\sigma^2 I$, that is
constant variance ($\sigma^2$) in the diagonal of the $6\times6$ square
matrix and 0's in all the off-diagonal cells:

```{r}
sigma <- 2
SigmaK <- diag(sigma^2, 6, 6)
dimnames(SigmaK) <- list(rownames(M), rownames(M))
print_tb(SigmaK)
```

We can use the `edma_simulate_data` function to simulate `n`
specimen given form matrix `M` and covariance matrix `SigmaK`:

```{r}
n <- 1000
sim <- edma_simulate_data(n=n, M, SigmaK)

tmp <- plot_2d(sim, xlim=c(-80, 100), ylim=c(-60, 60), asp=1)
xy <- attr(tmp, "coordinates")
text(xy[,1]+15, xy[,2], labels=rownames(xy), cex=0.6, col=4)
```

We can see from the figure that the constant variance translates to
roughly circle shaped convex hulls around the landmarks
(the hull bounds the landmarks form the `n` specimens),
and the size of these patches is similar (equal variances).

## Nonparametric estimation

We use the non-parametric EDMA algorithm to estimate
the form matrix and the SigmaKstar matrix:

```{r}
fit <- edma_fit(sim)
```

Because we used simulations, we can compare the known `M` matrix
with the estimated mean form:

```{r}
M                       # true
round(Meanform(fit), 2) # estimate
```

The 'true' `SigmaKstar` matrix is stored in our simulation object:

```{r}
tmp <- attr(sim, "simulation_settings")$SigmaKstar
round(tmp, 2)
round(SigmaKstar(fit), 2)
```

## Pattern matrix

The parametric estimation requires a pattern matrix
that describes the unknowns in the `SigmaK` matrix.
This `SigmaK` matrix is the variance covariance matrix
describing the variances (diagonal) associated with each landmark,
and the covariances (off-diagonal) among the landmarks.

The pattern matrix can be constructed in R, but when the number
of landmarks is high and variances are expected to differ among
the landmarks, it might be easier to construct the pattern matrix
in a spreadsheet editor (Excel). Here is how the file should be
constructed (the package contains example files, see
`?read_pattern`):

- the 1st row describes the landmarks starting from the 2nd column,
- the 1st column describes the landmarks in the same order as the column headers,
- leave the cell in the top left corner empty,
- use unique names or numbers in the rest of the table to denote unique variables in the Sigma matrix,
- structural zeros (cells where the covariance is known or assumed to be 0) have to be blank,
- there cannot be blank cells in the diagonal of the matrix,
- values below and above the diagonal must be symmetric (value in row 2/column 3 must be the same as row 3/column 2), either unique names or numbers, or blank.

The pattern matrix for the equal variance case looks like this:

```{r}
p <- matrix(
  c("1", NA, NA, NA,  NA,  NA,
    NA, "1", NA, NA,  NA,  NA,
    NA, NA, "1", NA,  NA,  NA,
    NA,  NA,  NA, "1", NA, NA,
    NA,  NA,  NA, NA, "1", NA,
    NA,  NA,  NA, NA, NA, "1"),
  nrow=6, ncol=6, byrow=TRUE)
dimnames(p) <- list(rownames(M), rownames(M))
print_tb(p)
```

## Parametric estimation

The parametric estimation requires the fitted object from
the previous step of nonparametric estimation
(that gave us the estimate of the mean form and SigmaKstar),
and the pattern matrix, so that the program knows what parameters
to estimate and where those are placed in the SigmaK matrix:

```{r}
(o <- SigmaK_fit(fit, p))
```

Now let us compare the 'true' SigmaK matrix with the estimated ones.
The 'full' (unconstrained) estimate can be accessed by the
`SigmaKfull` function which follows Hu (2007).
The `SigmaK` returns the estimates that are constrained by the
pattern matrix. This result makes sure that the parameters coded 
with the same names in the pattern matrix `p` get the same 
estimated parameter values. This is achived by
an additional optimization step based on the unconstrained solution.

```{r}
print_tb(SigmaK)              # true
print_tb(round(SigmaKfull(o), 2)) # full estimate
print_tb(round(SigmaK(o), 2)) # constrained estimate
cbind(true=parm, est=o$results$par[names(parm)])
```

## Unequal variance case

In the unequal variance case we can have different variances for the
different landmarks. Here is the corresponding pattern matrix:

```{r}
p <- matrix(
  c("1", NA,  NA, NA,  NA,  NA,
    NA, "1", NA, NA,  NA,  NA,
    NA,  NA, "1", NA,  NA,  NA,
    NA,  NA,  NA, "2", NA,  NA,
    NA,  NA,  NA, NA, "2", NA,
    NA,  NA,  NA, NA,  NA, "2"),
  nrow=6, ncol=6, byrow=TRUE)
dimnames(p) <- list(rownames(M), rownames(M))
plot_tb(p)
```

The SigmaK matrix looks like this:

```{r}
parm <- c("1"=12, "2"=8)
SigmaK <- make_Sigma(parm, p)
dimnames(SigmaK) <- dimnames(p)
print_tb(SigmaK)
```

The unequal variances translate to the following simulated data:

```{r}
sim <- edma_simulate_data(n=n, M, SigmaK)

tmp <- plot_2d(sim, xlim=c(-80, 100), ylim=c(-60, 60), asp=1)
xy <- attr(tmp, "coordinates")
text(xy[,1]+15, xy[,2], labels=rownames(xy), cex=0.6, col=4)
```

We can see the different sizes of the convex hulls corresponding to the
variances for each landmark.

For the estimation, we use `edma_fit` and `SigmaK_fit` as before:

```{r}
fit <- edma_fit(sim)
o <- SigmaK_fit(fit, p)
print_tb(SigmaK)              # true
print_tb(round(SigmaKfull(o), 2)) # full estimate
print_tb(round(SigmaK(o), 2)) # constrained estimate
cbind(true=parm, est=o$results$par[names(parm)])
```

Let's have different values in the diagonal:

```{r}
p <- matrix(
  c("1", NA,  NA, NA,  NA,  NA,
    NA, "2", NA, NA,  NA,  NA,
    NA,  NA, "3", NA,  NA,  NA,
    NA,  NA,  NA, "4", NA,  NA,
    NA,  NA,  NA, NA, "5", NA,
    NA,  NA,  NA, NA,  NA, "6"),
  nrow=6, ncol=6, byrow=TRUE)
dimnames(p) <- list(rownames(M), rownames(M))
plot_tb(p)
```

The SigmaK matrix looks like this:

```{r}
parm <- c("1"=12, "2"=8, "3"=11, "4"=6, "5"=10, "6"=4)
SigmaK <- make_Sigma(parm, p)
dimnames(SigmaK) <- dimnames(p)
print_tb(SigmaK)
```

This translates to the following simulated data:

```{r}
sim <- edma_simulate_data(n=n, M, SigmaK)

tmp <- plot_2d(sim, xlim=c(-80, 100), ylim=c(-60, 60), asp=1)
xy <- attr(tmp, "coordinates")
text(xy[,1]+15, xy[,2], labels=rownames(xy), cex=0.6, col=4)
```


We can now see that the constrained and full solutions are identical,
because no parameters are repeated in our pattern matrix:

```{r}
fit <- edma_fit(sim)
o <- SigmaK_fit(fit, p)
print_tb(SigmaK)              # true
print_tb(round(SigmaKfull(o), 2)) # full estimate
print_tb(round(SigmaK(o), 2)) # constrained estimate
cbind(true=parm, est=o$results$par[names(parm)])
```


## Correlations

We can add correlations between the frontal (L1) and lateral (L2, L3)
landmarks, and similarly to the other 3 landmarks:

```{r}
p <- matrix(
  c("1", "2", "2", NA,  NA,  NA,
    "2", "1", NA,  NA,  NA,  NA,
    "2", NA,  "1", NA,  NA,  NA,
    NA,  NA,  NA,  "3", "4", "4",
    NA,  NA,  NA,  "4", "3", NA,
    NA,  NA,  NA,  "4", NA,  "3"),
  nrow=6, ncol=6, byrow=TRUE)
dimnames(p) <- list(rownames(M), rownames(M))
plot_tb(p)
```

The SigmaK matrix looks like this:

```{r}
parm <- c("1"=12, "2"=1, "3"=8, "4"=2)
SigmaK <- EDMAinR:::.vec2mat(parm, EDMAinR:::.mat2fac(p))
dimnames(SigmaK) <- dimnames(p)
print_tb(SigmaK)
```

The unequal variances translate to the following simulated data:

```{r}
sim <- edma_simulate_data(n=n, M, SigmaK)

tmp <- plot_2d(sim, xlim=c(-80, 100), ylim=c(-60, 60), asp=1)
xy <- attr(tmp, "coordinates")
text(xy[,1]+15, xy[,2], labels=rownames(xy), cex=0.6, col=4)
```

This setup seems estimable:

```{r}
fit <- edma_fit(sim)
o <- SigmaK_fit(fit, p)
print_tb(SigmaK)              # true
print_tb(round(SigmaK(o), 2)) # estimate
```

## Non-estimable case: equi-correlation

The following parametrization is non identifiable:

```{r}
p <- matrix(
  c("1", "2", "2", "2", "2", "2",
    "2", "1", "2", "2", "2", "2",
    "2", "2", "1", "2", "2", "2",
    "2", "2", "2", "1", "2", "2",
    "2", "2", "2", "2", "1", "2",
    "2", "2", "2", "2", "2", "1"),
  nrow=6, ncol=6, byrow=TRUE)
dimnames(p) <- list(rownames(M), rownames(M))
plot_tb(p)
```

The diagonal elements are $\sigma^2$ (equal variance), the off diagonal
elements are $\sigma^2\rho$ where $\rho$ is the correlation:

```{r}
parm <- c("1"=10, "2"=2)
SigmaK <- EDMAinR:::.vec2mat(parm, EDMAinR:::.mat2fac(p))
dimnames(SigmaK) <- dimnames(p)
print_tb(SigmaK)
```

The correlation lead to more ellipsoid shapes for the simulated landmarks:

```{r}
sim <- edma_simulate_data(n=n, M, SigmaK)

tmp <- plot_2d(sim, xlim=c(-80, 100), ylim=c(-60, 60), asp=1)
xy <- attr(tmp, "coordinates")
text(xy[,1]+15, xy[,2], labels=rownames(xy), cex=0.6, col=4)
```

```{r}
fit <- edma_fit(sim)
try(o <- SigmaK_fit(fit, p))
```


## Non-estimable case: blocks

We can add correlations between the frontal (L1) and lateral (L2, L3)
landmarks, and similarly to the other 3 landmarks:

```{r}
p <- matrix(
  c("1", "2", "2", NA,  NA,  NA,
    "2", "1", "2",  NA,  NA,  NA,
    "2", "2", "1", NA,  NA,  NA,
    NA,  NA,  NA,  "3", "4", "4",
    NA,  NA,  NA,  "4", "3", "4",
    NA,  NA,  NA,  "4", "4","3"),
  nrow=6, ncol=6, byrow=TRUE)
dimnames(p) <- list(rownames(M), rownames(M))
plot_tb(p)
```

The SigmaK matrix looks like this:

```{r}
parm <- c("1"=12, "2"=1, "3"=8, "4"=2)
SigmaK <- EDMAinR:::.vec2mat(parm, EDMAinR:::.mat2fac(p))
dimnames(SigmaK) <- dimnames(p)
print_tb(SigmaK)
```

The unequal variances translate to the following simulated data:

```{r}
sim <- edma_simulate_data(n=n, M, SigmaK)

tmp <- plot_2d(sim, xlim=c(-80, 100), ylim=c(-60, 60), asp=1)
xy <- attr(tmp, "coordinates")
text(xy[,1]+15, xy[,2], labels=rownames(xy), cex=0.6, col=4)
```

This setup seems estimable:

```{r}
fit <- edma_fit(sim)
try(o <- SigmaK_fit(fit, p))
```


