---
title: "EDMA data"
author: "Peter Solymos"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EDMA data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup,include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(429)
```

## Introduction

**EDMAinR** is an R package for Euclidean Distance Matrix Analysis (EDMA).
EDMA is a coordinate‐free approach for comparing biological shapes using landmark data.

We can install the package from GitHub as:

```{r install,eval=FALSE}
if (!require(EDMAinR)) {
  if (!require(remotes))
      install.packages("remotes")
  remotes::install_github("psolymos/EDMAinR")
}
```

We can now load the package:

```{r load}
library(EDMAinR)
```

## Landmark data

Use the `read_xyz` function to read 2 or 3 D landmark data from `*.xyz` files.
First we specify the paths to two xyz files:

```{r filexyz}
file <- system.file("extdata/crouzon/Crouzon_P0_Global_MUT.xyz", 
    package="EDMAinR")
```

Note: we use the `system.file()` function to access example files
from the package. When defining your own files,
you will either set the working directory using `setwd()` or
a path like `c:/Users/<user>/<etc>`.

Now we can read in these text files:

```{r readxyz}
x <- read_xyz(file)
x
```

The data objects are lists with 2 elements: 

- `$name` contains info about the file from its header
- `$data` contains the landmark data

Here are the methods that we can use to learn more about the data sets.

Access dimensions (landmarks, $K$; dimensions, $D$; replicates, $n$) 
and dimension names (`landmark_names` returns the landmark labels):

```{r datamethods}
dim(x)
dimnames(x)
landmarks(x)
dimensions(x)
specimens(x)
```

Subsetting the data comes handy sometimes. The most general way to subset
the data sets is via the `[` function, the 3 indices inside the brackets
refer to the landmarks, dimensions, and specimens:

```{r datasubset}
x[1:10, , ]  # select the 1st 10 landmarks
x[ , 1:2, ]  # select 2 of the 2 dimensions
x[ , , 1:20] # select the 1st 20 individuals
x[1:10, , 1:20] # combine multiple indices
```

The data (`$data`) format inside the object `x` is list of the $K \times D$
matrices for each individual. Sometimes it is handy to stack these
matrices and create a rectangular data (either as a matrix, or data frame,
with $n \times K$ rows and $D$ columns):

```{r dataflat}
str(as.matrix(x))
```

Note: we are using `str` to show the structure of these objects,
this is not necessary when exploring the data.

Alternatively, we can store the data as an array
($K \times D \times n$):

```{r dataarray}
str(as.array(x))
```

Data sets can be checked using the `plot` function. 
The function steps through all the specimens to check
for any surprises.
The grey areas shows the convex hull for the landmark
leaving the specimen out, while red dots show the
actual specimen.
We will insert an error to show how it can be helpful:

```{r fig.width=7,fig.height=5,out.width='45%',fig.show='hold'}
## original values for specimen 1
plot(x, which=1)
## we change the 1st landmark
xwrong <- x
xwrong$data[[1]][1,1:2] <- c(2, 2)
plot(xwrong, which=1)
```

The ordination and cluster dendrogram

```{r fig.width=7,fig.height=5,out.width='45%',fig.show='hold'}
plot_ord(x)
plot_ord(xwrong)
```

```{r fig.width=7,fig.height=7,out.width='45%',fig.show='hold'}
plot_clust(x)
plot_clust(xwrong)
```

