---
title: "EDMA data"
author: "Peter Solymos"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EDMA data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup,include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(429)
```

## Introduction

**EDMAinR** is an R package for Euclidean Distance Matrix Analysis (EDMA).
EDMA is a coordinate‐free approach for comparing biological shapes using landmark data.

We can install the package from GitHub as:

```{r install,eval=FALSE}
if (!require(EDMAinR)) {
  if (!require(remotes))
      install.packages("remotes")
  remotes::install_github("psolymos/EDMAinR")
}
```

We can now load the package:

```{r load}
library(EDMAinR)
```

## Landmark data

### Reading data from xyz files

Use the `read_xyz` function to read 2D or 3D landmark data from `*.xyz` files.
First we specify the paths to two xyz files:

```{r filexyz}
file <- system.file("extdata/crouzon/Crouzon_P0_Global_MUT.xyz", 
    package="EDMAinR")
```

> *Note*: we use the `system.file()` function to access example files
> from the package. When defining your own files,
> you will either set the working directory using `setwd()` or use
> a path like `c:/Users/<user>/<etc>`.

The xyz landmark data has the following structure:

- Header: this is the description of the data.
- `XYZ`: indicates dimensions, XYZ means 3D landmark data.
- `42L 3 9`: dimensions, e.g. 42 landmarks ($K$), 3 dimensions ($D$), 9 specimens ($n$).
- Landmark names, separated by space.
- The stacked data of landmark coordinates, e.g. 3 columns, space separated numeric values with $K \times n$ rows, the $K$ landmarks per individuals stacked $n$ times.
- Blank line.
- Date on of scans for each specimen ($n$ rows), this part is also used to get specimen IDs.

Now we can read in these text files:

```{r readxyz}
x <- read_xyz(file)
x
```

### The EDMA data format

The printout of `x` told us the most important facts about the data:
the header information and the dimensions.

The EDMA data object (class 'edma_data') is a list with two
elements: 

- `$name` is the data set name (header information from
the `.xyz` file), 
- `$data` is a list of $n$ matrices (the list can be named if speciemen information is present), each matrix is of dimension $K \times D$, dimension names for the matrices describing landmark names and coordinate names.

### EDMA data accessor functions

The following methods are used to retrieve useful information
contained in EDMA data sets.

Access dimensions (landmarks, $K$; dimensions, $D$; replicates, $n$) 
and dimension names (`landmarks` returns the landmark labels):

```{r datamethods}
dim(x) # dimensions
dimnames(x) # dimension names
landmarks(x) # landmark names
dimensions(x) # the names of the coordinate dimensions
specimens(x) # specimen IDs
```

When the dimension names are missing from the input data file,
we use `L` to denote landmarks and `S` to denote specimens. These are
prepended to numeric landmark and specimen IDs. The coordinate dimensions
are `"X"`, `"Y"`, and optionally `"Z"` for 3D landmark data. 

### Selecting subsets of EDMA data objects

Subsetting the data comes handy sometimes. The most general way to subset
the data sets is via the `[` function, the 3 indices inside the brackets
refer to the landmarks, dimensions, and specimens.
Imagine the data as a 3D data cube and the square brackets allow us to
slice the cube from 3 different directions.

```{r datasubset}
x[1:10, , ]  # select the 1st 10 landmarks
x[ , 1:2, ]  # select 2 of the 2 dimensions
x[ , , 1:20] # select the 1st 20 individuals
x[1:10, , 1:20] # combine multiple indices
```

Read more about how to use square brackets to extract
parts of the data object: `?Extract` (see section 'Matrices and arrays').

### Coercing EDMA data into other formats

The data (`$data`) format inside the object `x` is list of the $K \times D$
matrices for each individual. Sometimes it is handy to stack these
matrices and create a rectangular data (either as a matrix, or data frame,
with $n \times K$ rows and $D$ columns):

```{r dataflat}
str(as.matrix(x))
```

> *Note*: we are using `str` to show the structure of these objects,
> this is not necessary when exploring the data.

Alternatively, we can store the data as an array
($K \times D \times n$):

```{r dataarray}
str(as.array(x))
```

### Turning 3D arrays into EDMA data objects

The `as.edma_data` function turns a 3D array to an EDMA data object.
This is useful when handling 3D array objects returned by many
functions of the [**geomorph**](https://cran.r-project.org/package=geomorph) 
package (i.e. after reding Morphologika, NTS, TPS files).

The header information gets lost in this way, but we can set it quite easily:

```{r arraytoedma}
xnew <- as.edma_data(as.array(x))
xnew
xnew$name <- "This is the same data as before"
xnew
```

## Visual inspection of the data

The `plot_2d` function shows the convex hulls or confidence ellipses 
for the landmarks based on all the specimens.
The plot is showing the first two axes of multidimensional scaling 
based on the form matrices (FM; pairwise Eiclidean distances):

```{r allspec}
plot_2d(x)
```

The images by default show the convex hull for each landmark
enclosing all specimens.
It is also possible to use 95% confidence ellipses:

```{r allspec2}
plot_2d(x, hull=FALSE)
```

### Spotting data errors

Data sets can be checked using the `plot` function.
By default, this function steps through all the specimens to check for any surprises.
The blue areas shows the convex hull for the landmark
leaving a single specimen out, while red dots show the actual specimen.

Using the `which` argument we can specify which specimen we want to
display (it is not possible to select multiple specimen, use the `[` method
to take a subset of the data that is being plotted):

```{r onespec}
plot(x, which=1)
```

We will insert an error to show how the `plot` function can help identify problems.
We change the coordinates for the 1st landmark of the 1st specimen.
The error is showing up as an outlier (long red line):

```{r fig.width=7,fig.height=5,out.width='45%',fig.show='hold'}
## original values for specimen 1
plot(x, which=1)
## we change the 1st landmark
xwrong <- x
xwrong$data[[1]][1,1:2] <- c(2, 2)
plot(xwrong, which=1)
```

### Ordination and clustering

The ordination plot shows the specimens based on multidimensional scaling
using pairwise distances between specimens.
These distances are based on the T-statistic and can be calculated as
`as.dist(x)`. The calculation involves the following steps:

- calculate the form difference matrix (FDM) for speciem $i$ and $j$: $FDM_{ij} = FM_{i} / FM_{j}$ ($i, j = 1, \dots, n$ ),
- calculate the T-statistic on the log scale (so that a max/min ration of 1, that is 0 distance, becomes log(1)=0): $log(max(FDM_{ij}) / min(FDM_{ij}))$.

The multidimensional scaling uses the square root transformed log-T distances to make
distances more Euclidean.

The error shows up nicely in the ordination plot (colored red):

```{r fig.width=7,fig.height=5,out.width='45%',fig.show='hold'}
plot_ord(x, col=c(2, rep(1, dim(x)[3])))
plot_ord(xwrong, col=c(2, rep(1, dim(xwrong)[3])))
```

The cluster dendrogram is based on the same log-T distance matrix described above.
The agglomeration algorithm is set to the Ward's minimum variance method (`"ward.D2")
which is using the square root of the log-T distances.

```{r fig.width=7,fig.height=7,out.width='45%',fig.show='hold'}
plot_clust(x)
plot_clust(xwrong)
```

Other agglomeration methods can be set by passing a different `method` argument,
e.g. single linkage (see `?hclust` for a list of options):

```{r fig.width=7,fig.height=7,out.width='45%',fig.show='hold'}
plot_clust(x, method = "single")
plot_clust(xwrong, method = "single")
```

Both the ordination and clustering functions return the ordination/cluster
results, so that plots can be forther customized (see `?ape::plot.phylo`):

```{r}
mds <- plot_ord(x, plot=FALSE)
str(mds)
plot(mds$points, pch=4)

h <- plot_clust(x, plot=FALSE)
h
plot(h)
plot(ape::as.phylo(h), type="cladogram", font=1, cex=0.6)
```

