---
title: "Parametric estimation"
author: "Peter Solymos"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parametric estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup,include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(123)
```

## Introduction

This tutorial describes how to estimate SigmaK and assess if it is estimable.
Parametric estimation of SigmaK involves the following steps:

1. non-parametric estimation of mean form and SigmaKstar using the `edma_fit` function,
2. constructing of a pattern matrix based on biology of landmarks and respecting statistical constraints,
3. use the `SigmaK_fit` to estimate SigmaK using the non-parametric fit and the pattern matrix.

```{r}
library(EDMAinR)
```

## Simulated landmark data

To demonstrate the workflow, let us simulate some data.
Simulation is useful for checking the limits of the estimating procedure.
We know what the 'true' input are, because we are controlling those
through the simulation. We can then compare the outputs (our estimates),
to the known inputs, and this is how we evaluate the utility 
and limitations of our methods.

Let us begin with the following set of 6 landmarks in 2-dimensions:

```{r}
(M <- array(
  c(82, 21, 22, -50, -37, -37, 
  0,  17, -17, 0,  41,  -41), 
  dim=c(6, 2),
  dimnames=list(paste0("L", 1:6), c("X", "Y"))))
plot(M, pch=3, asp=1, axes=FALSE, ann=FALSE, type="n")
text(M, labels=rownames(M), col=4)
```

The simplest SigmaK variance-covariance matrix is $\sigma^2 I$, that is
constant variance ($\sigma^2$) in the diagonal of the $6\times6$ square
matrix and 0's in all the off-diagonal cells:

```{r}
sigma <- 2
SigmaK <- diag(sigma^2, 6, 6)
dimnames(SigmaK) <- list(rownames(M), rownames(M))
print_tb(SigmaK)
```

We can use the `edma_simulate_data` function to simulate `n`
specimen given form matrix `M` and covariance matrix `SigmaK`:

```{r}
n <- 1000
sim <- edma_simulate_data(n=n, M, SigmaK)

tmp <- plot_2d(sim, xlim=c(-80, 100), ylim=c(-60, 60), asp=1)
xy <- attr(tmp, "coordinates")
text(xy[,1]+15, xy[,2], labels=rownames(xy), cex=0.6, col=4)
```

We can see from the figure that the constant variance translates to
roughly circle shaped convex hulls around the landmarks
(the hull bounds the landmarks form the `n` specimens),
and the size of these patches is similar (equal variances).

## Nonparametric estimation

We use the non-parametric EDMA algorithm to estimate
the form matrix and the SigmaKstar matrix:

```{r}
fit <- edma_fit(sim)
```

Because we used simulations, we can compare the known `M` matrix
with the estimated mean form:

```{r}
M                       # true
round(Meanform(fit), 2) # estimate
```

The 'true' `SigmaKstar` matrix is stored in our simulation object:

```{r}
round(attr(sim, "simulation_settings")$SigmaKstar, 2)
round(SigmaKstar(fit), 2)
```

## Pattern matrix

The parametric estimation requires a pattern matrix
that describes the unknowns in the `SigmaK` matrix.
This `SigmaK` matrix is the variance covariance matrix
describing the variances (diagonal) associated with each landmark,
and the covariances (off-diagonal) among the landmarks.

The pattern matrix can be constructed in R, but when the number
of landmarks is high and variances are expected to differ among
the landmarks, it might be easier to construct the pattern matrix
in a spreadsheet editor (Excel). Here is how the file should be
constructed (the package contains example files, see
`?read_pattern`):

- the 1st row describes the landmarks starting from the 2nd column,
- the 1st column describes the landmarks in the same order as the column headers,
- leave the cell in the top left corner empty,
- use unique names or numbers in the rest of the table to denote unique variables in the Sigma matrix,
- structural zeros (cells where the covariance is known or assumed to be 0) have to be blank,
- there cannot be blank cells in the diagonal of the matrix,
- values below and above the diagonal must be symmetric (value in row 2/column 3 must be the same as row 3/column 2), either unique names or numbers, or blank.

The pattern matrix for the equal variance case looks like this:

```{r}
p <- matrix(
  c("1", NA, NA, NA,  NA,  NA,
    NA, "1", NA, NA,  NA,  NA,
    NA, NA, "1", NA,  NA,  NA,
    NA,  NA,  NA, "1", NA, NA,
    NA,  NA,  NA, NA, "1", NA,
    NA,  NA,  NA, NA, NA, "1"),
  nrow=6, ncol=6, byrow=TRUE)
dimnames(p) <- list(rownames(M), rownames(M))
print_tb(p)
```

## Parametric estimation

The parametric estimation requires the fitted object from
the previous step of nonparametric estimation
(that gave us the estimate of the mean form and SigmaKstar),
and the pattern matrix, so that the program knows what parameters
to estimate and where those are placed in the SigmaK matrix:

```{r}
(o <- SigmaK_fit(fit, p))
```

Now let us compare the 'true' SigmaK matrix with the estimated ones.
The 'full' estimate of SigmaK can be accessed by the
`SigmaKfull` function which follows Hu (2007).

```{r}
print_tb(SigmaK)                  # true
print_tb(round(SigmaKfull(o), 2)) # full estimate
```

We can see in the SigmaKfull matrix that the values are slightly
different, because the algorithm treats every non-`NA` entry in the
pattern matrix (non-0 cell in the SigmaK matrix to be estimated)
as an unknown variable.

The `SigmaK` by default returns the estimates that are constrained by the
pattern matrix. We call this default method the 2-step algorithm
because the numerical optimization follows the estimation of SigmaKfull.
We then use the pattern matrix to constrain the parameters denoted by the
same labels to be the same value. Let's see what value we get
for our constant variance parameter:

```{r}
print_tb(SigmaK)              # true
print_tb(round(SigmaK(o), 2)) # constrained estimate
cbind(true=sigma^2, estimate=o$results$par)
```

There is an other method we call the 1-step algorithm where the 
numerical optimization is performed without the information
from SigmaKfull. It uses only the SigmaKstar matrix.
We can use this algorithm as:

```{r}
(o1 <- SigmaK_fit(fit, p, twostep=FALSE))
cbind(true=sigma^2, onestep=o1$results$par, twostep=o$results$par)
```

Both estimates look the same and very close to the true estimate.

## Unequal variance case

In the unequal variance case we can have different variances for the
different landmarks. Here is the corresponding pattern matrix:

```{r}
p <- matrix(
  c("1", NA,  NA, NA,  NA,  NA,
    NA, "1", NA, NA,  NA,  NA,
    NA,  NA, "1", NA,  NA,  NA,
    NA,  NA,  NA, "2", NA,  NA,
    NA,  NA,  NA, NA, "2", NA,
    NA,  NA,  NA, NA,  NA, "2"),
  nrow=6, ncol=6, byrow=TRUE)
dimnames(p) <- list(rownames(M), rownames(M))
plot_tb(p)
```

The SigmaK matrix looks like this:

```{r}
parm <- c("1"=12, "2"=2)
SigmaK <- make_Sigma(parm, p)
dimnames(SigmaK) <- dimnames(p)
print_tb(SigmaK)
```

The unequal variances translate to the following simulated data:

```{r}
sim <- edma_simulate_data(n=n, M, SigmaK)

tmp <- plot_2d(sim, xlim=c(-80, 100), ylim=c(-60, 60), asp=1)
xy <- attr(tmp, "coordinates")
text(xy[,1]+15, xy[,2], labels=rownames(xy), cex=0.6, col=4)
```

We can see the different sizes of the convex hulls corresponding to the
variances for each landmark.

For the estimation, we use `edma_fit` and `SigmaK_fit` as before:

```{r}
fit <- edma_fit(sim)
o1 <- SigmaK_fit(fit, p, twostep=FALSE)
o2 <- SigmaK_fit(fit, p, twostep=TRUE)
print_tb(SigmaK)                   # true
print_tb(round(SigmaKfull(o2), 2)) # full estimate
cbind(true=parm, 
  onestep=o1$results$par[names(parm)], 
  twostep=o2$results$par[names(parm)])
```

Let's have 6 different values in the diagonal:

```{r}
p <- matrix(
  c("1", NA,  NA, NA,  NA,  NA,
    NA, "2", NA, NA,  NA,  NA,
    NA,  NA, "3", NA,  NA,  NA,
    NA,  NA,  NA, "4", NA,  NA,
    NA,  NA,  NA, NA, "5", NA,
    NA,  NA,  NA, NA,  NA, "6"),
  nrow=6, ncol=6, byrow=TRUE)
dimnames(p) <- list(rownames(M), rownames(M))
plot_tb(p)
```

The SigmaK matrix looks like this:

```{r}
parm <- c("1"=12, "2"=8, "3"=11, "4"=3, "5"=10, "6"=2)
SigmaK <- make_Sigma(parm, p)
dimnames(SigmaK) <- dimnames(p)
print_tb(SigmaK)
```

This translates to the following simulated data:

```{r}
sim <- edma_simulate_data(n=n, M, SigmaK)

tmp <- plot_2d(sim, xlim=c(-80, 100), ylim=c(-60, 60), asp=1)
xy <- attr(tmp, "coordinates")
text(xy[,1]+15, xy[,2], labels=rownames(xy), cex=0.6, col=4)
```


We can now see that the constrained and full solutions are identical,
because no parameters are repeated in our pattern matrix:

```{r}
fit <- edma_fit(sim)
o1 <- SigmaK_fit(fit, p, twostep=FALSE)
o2 <- SigmaK_fit(fit, p, twostep=TRUE)
print_tb(SigmaK)                   # true
print_tb(round(SigmaKfull(o2), 2)) # full estimate
cbind(true=parm, 
  onestep=o1$results$par[names(parm)], 
  twostep=o2$results$par[names(parm)])
```

We can see that both the 1-step and 2-step algorithms give
estimates that are close to the true values when all the correlations (off-diagonal
elements) of SigmaK are 0 (coded as `NA` in the pattern matrix).

## Correlations

We can add correlations between the frontal (L1) and lateral (L2, L3)
landmarks, and similarly to the other 3 landmarks:

```{r}
p <- matrix(
  c("1", "2", "2", NA,  NA,  NA,
    "2", "1", NA,  NA,  NA,  NA,
    "2", NA,  "1", NA,  NA,  NA,
    NA,  NA,  NA,  "3", "4", "4",
    NA,  NA,  NA,  "4", "3", NA,
    NA,  NA,  NA,  "4", NA,  "3"),
  nrow=6, ncol=6, byrow=TRUE)
dimnames(p) <- list(rownames(M), rownames(M))
plot_tb(p)
```

The SigmaK matrix looks like this:

```{r}
parm <- c("1"=12, "2"=1, "3"=8, "4"=2)
SigmaK <- EDMAinR:::.vec2mat(parm, EDMAinR:::.mat2fac(p))
dimnames(SigmaK) <- dimnames(p)
print_tb(SigmaK)
```

The unequal variances translate to the following simulated data:

```{r}
sim <- edma_simulate_data(n=n, M, SigmaK)

tmp <- plot_2d(sim, xlim=c(-80, 100), ylim=c(-60, 60), asp=1)
xy <- attr(tmp, "coordinates")
text(xy[,1]+15, xy[,2], labels=rownames(xy), cex=0.6, col=4)
```

This setup seems estimable:

```{r}
fit <- edma_fit(sim)
o1 <- SigmaK_fit(fit, p, twostep=FALSE)
o2 <- SigmaK_fit(fit, p, twostep=TRUE)
print_tb(SigmaK)                   # true
print_tb(round(SigmaKfull(o2), 2)) # full estimate
cbind(true=parm, 
  onestep=o1$results$par[names(parm)], 
  twostep=o2$results$par[names(parm)])
```

## Sensitivity analysis

The parametric estimate is based on an optimization algorithm.
This algorithm requires starting values for the unknown parameters.
The robustness of our estimates can be tested using different
starting values for the optimization. The idea is that 
we can call the estimates stable if the optimization finds
the same values irrespective of the starting values.

We can assess the results in this regard using the `sensitivity`
function. This function repeats the estimation `m` times
by using a starting value that is the original starting value
times a random uniform number. The same idea can be applied to the
1-step and 1-step estimator:

```{r}
sensitivity(o1)
sensitivity(o2)
```

The `par_1`, `par_2`, etc. columns shows the parameter estimates for the
unknown variables, the `value` column shows the value of the loss 
function evaluated these parameter values. Each row represent
a separate run, the 1st row the original estimate, the rest
from repeated runs with different starting values.
We see that in this case the 2-step estimator is less 
sensitive to starting values, but both methods provide very similar
estimates and loss function values.

## Non-estimable case: equi-correlation

The following parametrization is a very common parametrization.
As we will see, contrary to the intuition that models with few
unknowns (a single variance and a single covariance term) are identifiable,
this model is actually non-identifiable. Let us walk through the steps
of estimating the parameters and look for signs of non-identifiability:

```{r}
p <- matrix(
  c("1", "2", "2", "2", "2", "2",
    "2", "1", "2", "2", "2", "2",
    "2", "2", "1", "2", "2", "2",
    "2", "2", "2", "1", "2", "2",
    "2", "2", "2", "2", "1", "2",
    "2", "2", "2", "2", "2", "1"),
  nrow=6, ncol=6, byrow=TRUE)
dimnames(p) <- list(rownames(M), rownames(M))
plot_tb(p)
```

The diagonal elements are $\sigma^2$ (equal variance), the off diagonal
elements are $\sigma^2\rho$ where $\rho$ is the correlation:

```{r}
parm <- c("1"=10, "2"=2)
SigmaK <- EDMAinR:::.vec2mat(parm, EDMAinR:::.mat2fac(p))
dimnames(SigmaK) <- dimnames(p)
print_tb(SigmaK)
```

The correlation lead to more ellipsoid shapes for the simulated landmarks:

```{r}
sim <- edma_simulate_data(n=n, M, SigmaK)

tmp <- plot_2d(sim, xlim=c(-80, 100), ylim=c(-60, 60), asp=1)
xy <- attr(tmp, "coordinates")
text(xy[,1]+15, xy[,2], labels=rownames(xy), cex=0.6, col=4)
```

```{r}
fit <- edma_fit(sim)
try(o1 <- SigmaK_fit(fit, p, twostep=FALSE))
try(o2 <- SigmaK_fit(fit, p, twostep=TRUE))
```

Oops, we get an error in both cases. Based on Hu (2007),
we need to have a certain number of 0s in the SigmaK matrix
to be able to identify the parameters. The number of non-zero
values cannot be larger than $K (K-1)/2$ (counting the elements
in the diagonal and the lower triangle only). This is what the 
error informs us about. The `check` argument can be used to
disable this check and see what we get if we disregard
such mathematical results:

```{r}
try(o1 <- SigmaK_fit(fit, p, twostep=FALSE, check=FALSE))
try(o2 <- SigmaK_fit(fit, p, twostep=TRUE, check=FALSE))
```

Interesting. The 1-step estimator works because it does not depend on
SigmaKfull. The 2-step estimator, however, fails because
the so called $A$ matrix that we need to estimate SigmaKfull
is non-invertible.

Let's investigate the sensitivity analysis results for the 1-step estimator:

```{r}
sensitivity(o1)
```

The clear sign of non-identifiability here is that the loss function values
are identical across runs, but the estimates vary widely.
Just as a reminder, our true values were `r parm["1"]` for variance
and `r parm["2"]` for covariance.

## Non-estimable case: blocks

We can add correlations between the frontal (L1) and lateral (L2, L3)
landmarks, and similarly to the other 3 landmarks:

```{r}
p <- matrix(
  c("1", "2", "2", NA,  NA,  NA,
    "2", "1", "2",  NA,  NA,  NA,
    "2", "2", "1", NA,  NA,  NA,
    NA,  NA,  NA,  "3", "4", "4",
    NA,  NA,  NA,  "4", "3", "4",
    NA,  NA,  NA,  "4", "4","3"),
  nrow=6, ncol=6, byrow=TRUE)
dimnames(p) <- list(rownames(M), rownames(M))
plot_tb(p)

parm <- c("1"=12, "2"=1, "3"=8, "4"=2)
SigmaK <- EDMAinR:::.vec2mat(parm, EDMAinR:::.mat2fac(p))
dimnames(SigmaK) <- dimnames(p)
print_tb(SigmaK)
```

The number of non-zero entries (diagonal and lower triangle)
in this case is 12 which is less than $K(K-1)/2=15$.
According to the simple rule of thumb from above, this should work.

The bloc design translates to the following simulated data:

```{r}
sim <- edma_simulate_data(n=n, M, SigmaK)

tmp <- plot_2d(sim, xlim=c(-80, 100), ylim=c(-60, 60), asp=1)
xy <- attr(tmp, "coordinates")
text(xy[,1]+15, xy[,2], labels=rownames(xy), cex=0.6, col=4)
```

Although 12 < 15, this setup still leads to non-ivertible $A$ matrix 
for the 2-step algorithm, and we can see the clear signs of non-estimability
for the 1-step algorithm based on the sensitivity analysis:

```{r}
fit <- edma_fit(sim)
try(o1 <- SigmaK_fit(fit, p, twostep=FALSE, check=FALSE))
try(o2 <- SigmaK_fit(fit, p, twostep=TRUE, check=FALSE))
sensitivity(o1)
```

## An estimable block pattern

Let's try a slight variation of this block design:

```{r}
p <- matrix(
  c("1", "2", "2", "2",  NA,  NA,
    "2", "1", NA,  NA,   NA,  NA,
    "2", NA,  "1", NA,   NA,  NA,
    "2", NA,  NA,  "3",  "4", "4",
    NA,  NA,  NA,  "4",  "3", "4",
    NA,  NA,  NA,  "4",  "4","3"),
  nrow=6, ncol=6, byrow=TRUE)
dimnames(p) <- list(rownames(M), rownames(M))
plot_tb(p)

parm <- c("1"=12, "2"=1, "3"=8, "4"=2)
SigmaK <- EDMAinR:::.vec2mat(parm, EDMAinR:::.mat2fac(p))
dimnames(SigmaK) <- dimnames(p)
print_tb(SigmaK)
```

We have again 12 non-zero entries, but the 0 from the 1st row and 2nd
colum to the 3rd row and 1st column (and its pair in the upper triangle).
```{r}
sim <- edma_simulate_data(n=n, M, SigmaK)

tmp <- plot_2d(sim, xlim=c(-80, 100), ylim=c(-60, 60), asp=1)
xy <- attr(tmp, "coordinates")
text(xy[,1]+15, xy[,2], labels=rownames(xy), cex=0.6, col=4)
```

Everything looks fine now:

```{r}
fit <- edma_fit(sim)
o1 <- SigmaK_fit(fit, p, twostep=FALSE)
o2 <- SigmaK_fit(fit, p, twostep=TRUE)

print_tb(SigmaK)                   # true
print_tb(round(SigmaKfull(o2), 2)) # full estimate
cbind(true=parm, 
  onestep=o1$results$par[names(parm)], 
  twostep=o2$results$par[names(parm)])

sensitivity(o1)
sensitivity(o2)
```

The default checks for the `SigmaK_fit` can catch the non-identifiable
situation. Use the `check=FALSE` and `twostep=FALSE` options with care.

## Selecting the best parametric model

Assumptions about biological mechanisms driving the 0 pattern in the
covariance matrix can help researchers to formulate hypotheses.
If these hypotheses translate into patterns that are estimable,
we can use the loss function value from the 1-step estimator
to tell which model fits the data better:

We will use the `p1` pattern matrix to simulate the data,
and we will use all 5 identifiable patterns to try to
estimate the parametric SigmaK matrix.

```{r}
p1 <- matrix(
  c("1", NA, NA, NA,  NA,  NA,
    NA, "1", NA, NA,  NA,  NA,
    NA, NA, "1", NA,  NA,  NA,
    NA,  NA,  NA, "1", NA, NA,
    NA,  NA,  NA, NA, "1", NA,
    NA,  NA,  NA, NA, NA, "1"),
  nrow=6, ncol=6, byrow=TRUE)
p2 <- matrix(
  c("1", NA,  NA, NA,  NA,  NA,
    NA, "1", NA, NA,  NA,  NA,
    NA,  NA, "1", NA,  NA,  NA,
    NA,  NA,  NA, "2", NA,  NA,
    NA,  NA,  NA, NA, "2", NA,
    NA,  NA,  NA, NA,  NA, "2"),
  nrow=6, ncol=6, byrow=TRUE)
p3 <- matrix(
  c("1", NA,  NA, NA,  NA,  NA,
    NA, "2", NA, NA,  NA,  NA,
    NA,  NA, "3", NA,  NA,  NA,
    NA,  NA,  NA, "4", NA,  NA,
    NA,  NA,  NA, NA, "5", NA,
    NA,  NA,  NA, NA,  NA, "6"),
  nrow=6, ncol=6, byrow=TRUE)
p4 <- matrix(
  c("1", "2", "2", NA,  NA,  NA,
    "2", "1", NA,  NA,  NA,  NA,
    "2", NA,  "1", NA,  NA,  NA,
    NA,  NA,  NA,  "3", "4", "4",
    NA,  NA,  NA,  "4", "3", NA,
    NA,  NA,  NA,  "4", NA,  "3"),
  nrow=6, ncol=6, byrow=TRUE)
p5 <- matrix(
  c("1", "2", "2", "2",  NA,  NA,
    "2", "1", NA,  NA,   NA,  NA,
    "2", NA,  "1", NA,   NA,  NA,
    "2", NA,  NA,  "3",  "4", "4",
    NA,  NA,  NA,  "4",  "3", "4",
    NA,  NA,  NA,  "4",  "4","3"),
  nrow=6, ncol=6, byrow=TRUE)
dimnames(p1) <- list(rownames(M), rownames(M))
dimnames(p2) <- list(rownames(M), rownames(M))
dimnames(p3) <- list(rownames(M), rownames(M))
dimnames(p4) <- list(rownames(M), rownames(M))
dimnames(p5) <- list(rownames(M), rownames(M))

parm <- c("1"=12, "2"=1, "3"=8, "4"=2)
SigmaK <- EDMAinR:::.vec2mat(parm, EDMAinR:::.mat2fac(p4))
dimnames(SigmaK) <- dimnames(p4)
sim <- edma_simulate_data(n=n, M, SigmaK)
```

We fit 5 models, then compare the loss function values.
The model with the smallest loss function value is the best:

```{r}
fit <- edma_fit(sim)
o1 <- SigmaK_fit(fit, p1, twostep=FALSE)
o2 <- SigmaK_fit(fit, p2, twostep=FALSE)
o3 <- SigmaK_fit(fit, p3, twostep=FALSE)
o4 <- SigmaK_fit(fit, p4, twostep=FALSE)
o5 <- SigmaK_fit(fit, p5, twostep=FALSE)

value <- sapply(list(o1, o2, o3, o4, o5), function(o) {
  o$results$value
})

data.frame(
  model=1:5,
  value=value,
  delta=value - min(value))
```
