---
title: "Parametric estimation"
author: "Peter Solymos"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parametric estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup,include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(429)
```

This section describes how to assess if SigmaK is esimable.

The pattern matrix describes the unknowns in the SigmaK matrix.
Diagonal must be filled. Upper and lower diagonal values must
be symmetric or `NA`. The file can be created in R,
or loaded as a csv or Excel (xls, xlsx) file:

```{r}
library(EDMAinR)

read_pattern(system.file("extdata/example.csv", package="EDMAinR"))
read_pattern(system.file("extdata/example.xlsx", package="EDMAinR"))
```

Now we go ahead with a simulated example:

```{r}
m <- matrix(c(
    "a", NA, NA, NA,
    NA, "a", NA, NA,
    NA,  NA, "b", NA,
    NA,  NA, NA, "b"
), 4, 4, byrow=TRUE)
parm <- c(a=0.25, b=0.35)

M <- structure(c(-2.5, 7.5, -2.5, -2.5, -7.5, 2.5, 2.5, 4.5),
    .Dim = c(4L, 2L))
SigmaK <- EDMAinR:::.vec2mat(parm, EDMAinR:::.mat2fac(m))

sim <- edma_simulate_data(n=500, M, SigmaK)
dimnames(M) <- dimnames(sim$data[[1L]])
rownames(SigmaK) <- rownames(m) <- rownames(sim$data[[1L]])
colnames(SigmaK) <- colnames(m) <- rownames(sim$data[[1L]])
```

This simulated example has 2 unknowns: `a` and `b`,
now we estimate these:

```{r}
fit <- edma_fit(sim)
o <- SigmaK_fit(fit, m)
o
cbind(true=parm, est=o$results$par)
SigmaK(o)
```

To check the sensitivity, we can re-estimate SigmaK matrix
multiple times with random starting values for numerical
optimization. When estimates vary while the loss function
value is the same, it is usually a sign of non-identifiability.

```{r}
s <- sensitivity(o)
summary(s)
boxplot(s)
```

Now let's try a non identifiable pattern.
The function is pretty smart to let us know, but we
will disable such checks, so that we can demonstrate
how do estimates and loss function value looks like in this case:
lots of different values can lead to the same value

```{r}
m <- matrix(c(
    "a", "b", "b", "b",
    "b", "a", "b", "b",
    "b", "b", "a", "b",
    "b", "b", "b", "a"
), 4, 4, byrow=TRUE)
parm <- c(a=0.25, b=0.07)


M <- structure(c(-2.5, 7.5, -2.5, -2.5, -7.5, 2.5, 2.5, 4.5),
    .Dim = c(4L, 2L))
SigmaK <- EDMAinR:::.vec2mat(parm, EDMAinR:::.mat2fac(m))

sim <- edma_simulate_data(n=500, M, SigmaK)
dimnames(M) <- dimnames(sim$data[[1L]])
rownames(SigmaK) <- rownames(m) <- rownames(sim$data[[1L]])
colnames(SigmaK) <- colnames(m) <- rownames(sim$data[[1L]])

fit <- edma_fit(sim)
try(SigmaK_fit(fit, m))

o <- SigmaK_fit(fit, m, check_pattern=FALSE)
o
cbind(true=parm, est=o$results$par)
SigmaK(o)

s <- sensitivity(o)
summary(s)
boxplot(s)
```
