---
title: "EDMA data"
author: "Peter Solymos"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EDMA data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup,include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(429)
```

# Introduction

**EDMAinR** is an R package for Euclidean Distance Matrix Analysis (EDMA).
EDMA is a coordinate‐free approach for comparing biological shapes using landmark data.

We can install the package from GitHub as with the help of the **remotes** R package:

```{r install,eval=FALSE}
if (!require(EDMAinR)) {
  if (!require(remotes))
      install.packages("remotes")
  remotes::install_github("psolymos/EDMAinR")
}
```

We can now load the **EDMAinR** package:

```{r load}
library(EDMAinR)
```

# Landmark data

## Reading data from xyz files

Use the `read_xyz` function to read 2D or 3D landmark data from `*.xyz` files.
First we specify the paths to two xyz files:

```{r filexyz}
file <- system.file("extdata/crouzon/Crouzon_P0_Global_MUT.xyz", 
    package="EDMAinR")
```

> *Note*: we use the `system.file()` function to access example files
> from the package. When defining your own files,
> you will either set the working directory using `setwd()` or use
> a path like `c:/Users/<user>/<etc>`.

The xyz landmark data has the following structure:

- Header: this is the description of the data.
- `XYZ`: indicates dimensions, XYZ means 3D landmark data.
- `42L 3 9`: dimensions, e.g. 42 landmarks ($K$), 3 dimensions ($D$), 9 specimens ($n$).
- Landmark names, separated by space.
- The stacked data of landmark coordinates, e.g. 3 columns, space separated numeric values with $K \times n$ rows, the $K$ landmarks per individuals stacked $n$ times.
- Blank line.
- Date on of scans for each specimen ($n$ rows), this part is also used to get specimen IDs.

Now we can read in these text files:

```{r readxyz}
x <- read_xyz(file)
x
```

> *Note*: find more information [below](#turning-3d-arrays-into-edma-data-objects) 
> about reading in other morphometrics file formats.

## The EDMA data format

The printout of `x` told us the most important facts about the data:
the header information and the dimensions.

The EDMA data object (class 'edma_data') is a list with two
elements: 

- `$name` is the data set name (header information from
the `.xyz` file), 
- `$data` is a list of $n$ matrices (the list can be named if specimen information is present), each matrix is of dimension $K \times D$, dimension names for the matrices describing landmark names (rows) and coordinate names (columns).

## EDMA data dimensions and names

The following methods are used to retrieve useful information
from EDMA data objects.

Use `dim` to access dimensions (landmarks, $K$; dimensions, $D$; replicates, $n$). 
`landmarks`, `dimansions`, and `specimens` return 
the landmark, dimension, and specimen names, respectively,
`dimnames` returns all three in a list:

```{r datamethods}
dim(x) # dimensions
dimnames(x) # dimension names
landmarks(x) # landmark names
dimensions(x) # the names of the coordinate dimensions
specimens(x) # specimen IDs
```

When the dimension names are missing from the input data file,
we use `L` to denote landmarks and `S` to denote specimens. These are
prepended to numeric landmark and specimen IDs. The coordinate dimensions
are `"X"`, `"Y"`, and optionally `"Z"` for 3D landmark data. 

It is also possible to manipulate dimension names.
This is useful when different files have the same landmarks
with slight changes (lower/upper case differences, etc.):


```{r datamethods2}
lm <- landmarks(x) # copy original names
landmarks(x) <- toupper(lm)
landmarks(x) # new landmark names
landmarks(x) <- lm # reset names to original
```


## Selecting subsets of EDMA data objects

Subsetting the data comes handy sometimes, i.e. we might want to
select only a few of the landmarks to get a simplified picture
of the relationships, or because different input files
list different landmarks and we want to select the ones that are
present in both data set.
Or we need to filter some specimens, because we want
to select only specimens of certain ages, gender, etc.
Often we need to draw a random sample of the specimens (for
randomization and bootstrap).

The most general way to subset
the data sets is *via* the `[` function (i.e. `x[i, j, k]`), 
the 3 indices inside the brackets
refer to the landmarks (`i`), dimensions (`j`), and specimens (`k`).
Imagine the data as a 3D data cube and the square brackets allow us to
slice the cube from 3 different directions.

```{r datasubset}
x[1:10, , ]  # select the 1st 10 landmarks
x[ , 1:2, ]  # select 2 of the 2 dimensions
x[ , , 11:20] # select individuals from 11 to 20
x[1:10, , 1:20] # combine multiple indices
```

Read more about how to use square brackets to extract
parts of the data object: `?Extract` (see section 'Matrices and arrays').
The most important thing to keep in mind is that the empty
index (e.g. `x[,,]`) means 'select everything'.

We can take subsets using:

- numeric indices (like as we saw above),
- logical indices,
- and by character names.

Logical indices might be handy if, e.g., we have ancillary data about
the specimens. If we have `age` of each specimens (values in ages matching
the order of specimens in the EDMA data object),
we can use it as `x[,,age < 1]`.

Character based subsets are useful if we want to select homologous
landmarks for multiple objects, or we want to make sure
that the landmarks' order is identical among multiple objects.
The same `[` function can be used not only to take a subset
but also to reorder landmarks or specimens. Character indices
are most suitable for this purpose.

## Coercing EDMA data into other object types

The data (`$data`) format inside the object `x` is list of the $K \times D$
matrices, one for each individual. Sometimes it is handy to stack these
matrices and create a rectangular data (either as a matrix, or data frame,
with $n \times K$ rows and $D$ columns):

```{r dataflat}
str(as.matrix(x))
```

> *Note*: we are using `str` to show the structure of these objects,
> this is not necessary when exploring the data.
> `str` here tells us the dimensions (`r nrow(as.matrix(x))` rows and
> `r ncol(as.matrix(x))` columns, the first few values, and the dimension names).

Alternatively, we can store the data as an array
($K \times D \times n$):

```{r dataarray}
str(as.array(x))
```

## Turning 3D arrays into EDMA data objects

The `as.edma_data` function turns a 3D array to an EDMA data object.
This is useful when handling 3D array objects returned by many
functions of the [**geomorph**](https://cran.r-project.org/package=geomorph) 
package (i.e. after reading Morphologika, NTS, TPS files).

The header information gets lost in this way, but we can set it quite easily:

```{r arraytoedma}
xnew <- as.edma_data(as.array(x))
xnew
xnew$name <- "This is the same data as before"
xnew
```

> *Note*: the **geomorph** package needs to be installed separately
> using `install.packages("geomorph")`.

# Visual inspection of the data

The `plot_2d` function shows the convex hulls or confidence ellipses 
for the landmarks based on all the specimens.
The plot is showing the first two axes of multidimensional scaling 
based on the form matrices ($FM$; pairwise Euclidean distances):

```{r allspec}
plot_2d(x)
```

The images by default show the convex hull for each landmark
enclosing all specimens.
It is also possible to use 95% confidence ellipses:

```{r allspec2}
plot_2d(x, hull=FALSE)
```

## Spotting data errors

Data sets can be checked using the `plot` function.
By default, this function steps through all the specimens to check for any surprises.
The blue areas show the convex hull for the landmark
leaving a single specimen out, while red dots show the actual specimen.
It is a sign of problem when a dot for a specimen is located far outside of 
the convex hull.

Using the `which` argument we can specify which specimen we want to
display (it is not possible to select multiple specimen, use the `[` method
to take a subset of the data that and plot it with `plot_2d`):

```{r onespec}
plot(x, which=1)
```

We will insert an error to show how the `plot` function can help identify problems.
We change the coordinates for the 1st landmark of the 1st specimen.
The error is showing up as an outlier (long red line):

```{r fig.width=7,fig.height=5,out.width='45%',fig.show='hold'}
## original values for specimen 1
plot(x, which=1)
## we change the 1st landmark
xwrong <- x
xwrong$data[[1]][1,1:2] <- c(2, 2)
plot(xwrong, which=1)
```

## Ordination and clustering

The ordination plot shows the specimens based on multidimensional scaling
using pairwise distances between specimens.
These distances are based on the T-statistic and can be calculated as
`as.dist(x)`. The calculation involves the following steps:

- calculate Euclidean distances ($ED_{i}$) between all pairs of landmarks for specimen $i$ ($i = 1, \dots, n$) based on the coordinates,
- calculate Euclidean distances ($ED_{j}$) between all pairs of landmarks for specimen $j$ ($j = 1, \dots, n$) based on the coordinates,
- calculate the ratio $ED_{i} / ED_{j}$,
- calculate the T-statistic on the log scale (so that a max/min ratio of 1, that is 0 distance, becomes log(1)=0): $d_{ij}=log(max[ED_{i} / ED_{j}] / min[ED_{i} / ED_{j}])$.

The multidimensional scaling uses the square root transformed log-T distances.

The error shows up nicely in the ordination plot (colored red):

```{r fig.width=7,fig.height=5,out.width='45%',fig.show='hold'}
plot_ord(x, col=c(2, rep(1, dim(x)[3])))
plot_ord(xwrong, col=c("red", rep("black", dim(xwrong)[3])))
```

The cluster dendrogram is based on the same log-T distance matrix described above.
The default agglomeration algorithm is set to the 
Ward's minimum variance method (`"ward.D2"`)
which is using the square root of the log-T distances too.

```{r fig.width=7,fig.height=7,out.width='45%',fig.show='hold'}
plot_clust(x)
plot_clust(xwrong)
```

Other agglomeration methods can be set by passing a different `method` argument,
e.g. single linkage (see `?hclust` for a list of options):

```{r fig.width=7,fig.height=7,out.width='45%',fig.show='hold'}
plot_clust(x, method = "single")
plot_clust(xwrong, method = "single")
```

Both the ordination and clustering functions return the ordination/cluster
results, so that plots can be further customized.
Here is an example to create a plot

```{r}
mds <- plot_ord(x, plot=FALSE)
str(mds)
plot(mds$points, pch=4, col="darkgreen",
  main="Ordination diagram")
abline(h=0, v=0, lty=2, col="grey")
text(mds$points, labels=specimens(x), pos=1, cex=0.6, col="tan")
```

[**ggplot2**](https://ggplot2-book.org/) is a popular alternative for
R's base graphics. Here is how to create plots with it:

```{r}
library(ggplot2)
df <- as.data.frame(mds$points)
p <- ggplot(data=df, aes(x=`Axis 1`, y=`Axis 2`, label=rownames(df))) +
  geom_label() +
  geom_vline(xintercept=0, lty=2) +
  geom_hline(yintercept=0, lty=2)
p
```

See `?ape::plot.phylo` for how to customize dendrograms:

```{r}
h <- plot_clust(x, plot=FALSE)
h
plot(h)
plot(ape::as.phylo(h), type="cladogram", font=1, cex=0.6)
```

[**ggdendro**](https://CRAN.R-project.org/package=ggdendro) is a package
that can plot tree-like data structures with **ggplot2**.
Here is an example:

```{r}
library(ggdendro)
ggdendrogram(h, rotate = TRUE, size = 2)
```
