---
title: "EDMA form difference matrix"
author: "Peter Solymos"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EDMA form difference matrix}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup,include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(429)
```

# Introduction

This tutorial explains how to calculate form difference based
on 2 EDMA data objects with homologous landmarks.
We will use 2 data sets: one with and the other without the
mutation responsible for Crouzon syndrome:

```{r}
library(EDMAinR)

file1 <- system.file("extdata/crouzon/Crouzon_P0_Global_MUT.xyz",
    package="EDMAinR")
x1 <- read_xyz(file1)

file2 <- system.file("extdata/crouzon/Crouzon_P0_Global_NON-MUT.xyz",
    package="EDMAinR")
x2 <- read_xyz(file2)
```

# Estimating the form matrices

We first estimate the mean forms (no bootstrap replicates are necessary).

```{r}
numerator <- edma_fit(x1, B=25)
denominator <- edma_fit(x2, B=25)
```

Form matrices ($FM$) are formed as pairwise Euclidean distances
between landmarks from EDMA fit objects using the estimated mean forms
from objects $A$ and $B$.

# Form difference

Form difference ($FDM$) is calculated as the ratio of form matrices
($FM$) from the numerator and denominator objects following
Lele and Richtsmeier (1992, 1995): $FDM(A,B) = FM(B)/FM(A)$.

Bootstrap distribution is based on fixing the
reference $FM$ and taking the ratio between the reference $FM$
and the bootstrap $FM$s from the other object.
The `ref_denom` argument can be used to control which object
is the reference (it is the denominator by default),
`B` is the number of replicates drawn from this distribution:

```{r}
fdm <- edma_fdm(numerator, denominator, B=25)
fdm
```

## Global T-test

The global T-test is based on the pairwise distances in the $FM$s, 
taking the max/min ratio of the distances.
This is then done for all the `B` replicates, which provides the null
distribution. 

Algorithm for global testing is as follows 
(suppose population 1 is the 'reference' population):

1. Resample $n_1$ observations from the **first** sample and compute $FM_1^*$.
2. Resample $n_2$ observations from the **first** sample and compute $FM_2^*$.
3. Compute the $FDM^* = FM_2^* / FM_1^*$ and $T^* = max(FDM^*) / min(FDM^*)$.
4. Repeat the above three steps `B` times to get the p-value.

`Tobs_test` provides a summary of the test, `plot_Ttest` helps to visualize
the null distribution (histogram) and the observed T statistic
(vertical line):

```{r fig.width=7,fig.height=5,out.width='100%'}
Tobs_test(fdm)
plot_Ttest(fdm)
```

## Local testing

The local testing is done based on the
confidence intervals using the stacked $FDM$s from the bootstrap.
Output structure is similar to the output of the `get_fm` function, but
the interpretation of the confidence intervals is different due to the
different nature of the bootstrap distribution.
Here the distribution characterizes the $FDM$ and not the $FM$.
The confidence level can be changed through the `level` argument.

The algorithm for confidence intervals for elements of the FDM is as follows:

1. Resample $n_1$ observations from the **first** sample and compute $FM_1^*$.
2. Resample $n_2$ observations from the **second** sample and compute $FM_2^*$.
3. Compute the $FDM* = FM_2^* / FM_1^*$.
4. Repeat the above three steps B times to get the confidence intervals for the elements of the $FDM$. 

```{r}
head(get_fdm(fdm))
head(get_fdm(fdm, sort=TRUE, decreasing=TRUE))
head(get_fdm(fdm, sort=TRUE, decreasing=FALSE))
```

Differences below or above the confidence limits
indicate distances that significantly deviate from the bootstrap
based null expectation, and thus are related to landmarks
that drive the differences. Inspecting the highest and lowest
differences (using `sort`) can help revealing these landmarks.

The `lower` and `upper` limits of the confidence intervals
are based on `confint` (row names indicate the unsorted sequence
of landmark pairs as in the output from `get_fdm(fdm)`):

```{r}
head(confint(fdm))
```

The `plot_ci` function shows the pairwise differences and confidence intervals.
The x-axis label shows the landmark pairs, and gets really busy.
Use the `xshow=FALSE` argument to remove labels.

```{r fig.width=7,fig.height=7,out.width='100%'}
plot_ci(fdm, xshow=FALSE)
```

We can use the stacked $FDM$ data frame to make a 
[**ggplot2**](https://ggplot2-book.org/)
based plot:

```{r fig.width=7,fig.height=5,out.width='100%'}
library(ggplot2)
df <- get_fdm(fdm, sort=TRUE)
df$x <- 1:nrow(df) # make x-axis continuous
df$name <- paste(df$row, df$col) # add names
p <- ggplot(data=df, aes(x=x, y=dist, 
    ymin=lower, ymax=upper, label=name)) +
  geom_ribbon(col="#0000ff40", fill="#0000ff40") +
  geom_line() +
  geom_hline(yintercept=1, col=2) +
  labs(y="FDM Ratio", x="Landmark Pairs") +
  theme_linedraw() +
  theme(axis.text.x=element_blank())
p
```

Once we have a **ggplot2** version of a plot, we can
load the **plotly** package to make the object interactive:

```{r message=FALSE}
library(plotly)
ggplotly(p)
```

# Influential landmarks

We can consider a landmark influential with respect to the form difference
if after removing the landmark, the global T value moves close to 1.
Such a case would indicate that the landmark in question is driving the
form differences (i.e. pairwise distances between the landmark and
other landmarks contribute to the maximum value in the T statistic).
If, however, we remove a non-influential landmark, 
we expect the T value not to change a lot. 
Therefore, the 'drop' in the T value after removing a landmark can be used
to rank landmarks based on their influence.

Influential landmarks are identified by leaving one landmark out at a time,
then calculating the the T-statistic based on the remaining distances.
We can use the bootstrap distribution to see of the T value 'drop'
makes the global test non-significant. This means that after
removing the landmark, the form difference cannot be distinguished
from the null expectation.

The most influential landmark is the one with the largest drop in T value
compared to the original T statistic.
`Tdrop` is the newly calculated T value after leaving out the landmark
in question:

```{r}
infl <- get_influence(fdm)
head(infl[order(infl$Tdrop),], 10)
```

> *Note*: we used the `order` function to create an index to order
> the rows of the `infl` data frame.
> `get_influence` also takes a `level` argument for specifying the
> desired confidence interval (default is 95%).

The `plot` function shows the landmarks ordered by `Tdrop`
with the influential landmarks on the left-hand side of the plot.
The horizontal line on the top indicates the original T value
(with all the landmarks considered), the increasing line
shows `Tdrop`, the shaded area is the null distribution for the
T statistic:

```{r fig.width=7,fig.height=5,out.width='100%'}
plot(infl)
```

Here is the **ggplot2** version:

```{r fig.width=7,fig.height=5,out.width='100%'}
df <- infl[order(infl$Tdrop),]
df$landmark <- factor(as.character(df$landmark), as.character(df$landmark))
p <- ggplot(data=df, aes(x=landmark, y=Tdrop, ymin=lower, ymax=upper, group=1)) +
  geom_ribbon(col="#0000ff40", fill="#0000ff40") +
  geom_hline(yintercept=Tobs_test(fdm)$statistic, col=2) +
  geom_line() +
  labs(y="T-value", x="Landmarks") +
  theme(axis.text.x=element_text(angle = 45,
        vjust = 1, hjust=1))
p
```

An the interactive version:

```{r}
ggplotly(p)
```

# Ordination and clustering for specimens

The ordination and cluster dendrogram shows the two sets
of specimens from the 2 objects in the same diagram.
The 2 sets are combined with the `combine_data` function:

```{r}
x1
x2
(x12 <- combine_data(x1, x2))
table(x12$groups)
```

The visualization otherwise use the same principles as
described for EDMA data objects. The difference is
that the specimens and their labels are colored
(using a the default qualitative palette) according to the
`groups` (numerator vs. denominator).

If the the numerator and denominator objects are different
(global T value is high, $p$ value is low, there are influential landmarks),
we expect the two groups to separate in ordination space
and along the dendrogram:

```{r fig.width=7,fig.height=7,out.width='100%'}
plot_ord(fdm)
```

The dendrogram leaves (specimen labels) are also colored by groups:

```{r fig.width=7,fig.height=7,out.width='100%'}
plot_clust(fdm)
```

The colors can be changed via the color options:

```{r fig.width=7,fig.height=7,out.width='100%'}
op <- options("edma_options" = list(
    diverging = "Blue-Red",
    qualitative = "Warm"))
plot_ord(fdm)
options(op)
```

# Visualizing landmarks

The 2D and 3D plots produce a plot of the mean form
from the reference object ('prototype').
The color intensity for the landmarks (dots) is associated with the `Tdrop`
influence value (larger the difference, the more intensive the color;
red by default).
Lines between the landmarks represent distances.
We use the diverging palettes:
<1 differences are colored blue (1st half of the palette), >1 differences 
are colored red (2st half of the palette).

```{r fig.width=7,fig.height=5,out.width='100%'}
plot_2d(fdm, cex=2)
```



```{r fig.width=7,fig.height=5,out.width='100%'}
library(rgl)
xyz <- plot_3d(fdm)
text3d(xyz, texts=rownames(xyz), pos=1) # this adds names
decorate3d() # this adds the axes
rglwidget(width = 600, height = 600, reuse = FALSE)
```

The 2D and 3D plots do not display all the pairwise distances.
If displaying all edges is desired, use the `all=TRUE` argument.



