---
title: "EDMAinR: basic usage"
output: rmarkdown::html_vignette
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{EDMAinR: basic usage}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r pre,include=FALSE}
#knitr::opts_chunk$set(eval=FALSE)
set.seed(1)
```

## Introduction

**EDMAinR** is an R package for Euclidean Distance Matrix Analysis (EDMA).
EDMA is a coordinate‐free approach for comparing biological shapes using landmark data
as described in Lele and Richtsmeier (1991).
The implementation follows Hu (2007).

## Install

```{r install,eval=FALSE}
if (!require(remotes))
    install.packages("remotes")
remotes::install_github("psolymos/EDMAinR")
```

We can now load the package:

```{r load}
library(EDMAinR)
```

## Landmark data

Use the `read_xyz` function to read 2 or 3 D landmark data from `*.xyz` files:

```{r readxyz}
file1 <- system.file("extdata/crouzon/Crouzon_P0_Global_MUT.xyz", 
    package="EDMAinR")
x1 <- read_xyz(file1)
x1

file2 <- system.file("extdata/crouzon/Crouzon_P0_Global_NON-MUT.xyz", 
    package="EDMAinR")
x2 <- read_xyz(file2)
x2
```

Here are the methods that we can use to learn more about the data sets.

Access dimensions (landmarks, $K$; dimensions, $D$; replicates, $n$) 
and dimension names (`landmark_names` returns the landmark labels):

```{r datamethods}
dim(x1)
dimnames(x1)
landmark_names(x1)
```

Subsetting the data comes handy sometimes. The most general way to subset
the data sets is via the `[` function, the 3 indices inside the brackets
refer to the landmarks, dimensions, and replicates (most often individuals).
The `subset` method subsets the replicates:

```{r datasubset}
x1[1:10, 2:3, 1:5]
subset(x1, 1:10)
```

The data format inside the objects `x1` and `x2` is list of the $K \times D$
matrices for each individual. Sometimes it is handy to stack these
matrices and create a rectangular data (either as a matrix, or data frame,
with $n \times K$ rows and $D$ columns):

```{r dataflat}
str(as.matrix(x1))
str(as.data.frame(x1))
str(stack(x1))
```

Alternatively, we can store the data as an array
($K \times D \times n$):

```{r dataarray}
str(as.array(x1))
```

## Nonparametric estimation

The nonparametric estimator gives the mean form matrix ($\hat{M}$)
and $\hat{\Sigma}_{K}^\ast$, that we can extract from the fitted
model object `fit` using the `Meanform` and `SigmaKstar` functions
(using only the first 5 landmarks here):

```{r fit}
fit <- edma_fit(x1[1:5,,])
fit
Meanform(fit)
SigmaKstar(fit)
```

We can extract the mean form as pairwise Euclidean distances
(object class _dist_ as it is customary in R, see `?dist` for the details).
This is the form matrix in distance matrix format:

```{r ed}
as.dist(fit)
```

The Euclidean distances from the form matrix can be stacked, 
and the stacked distances sorted. The `edma_fm` function
produces the stacked for matrix, because this is most useful to us
as it can be sorted and more easily inspected:

```{r stackeddist}
edma_fm(fit)
edma_fm(fit, sort=TRUE, decreasing=TRUE)
edma_fm(fit, sort=TRUE, decreasing=FALSE)
```

We can turn the Euclidean distance into principal components
using the `edma_pcs` function:

```{r pca}
pc <- edma_pca(fit)
plot(pc)
```



## Comparing 2 sets of 3D landmark data

We can fit the EDMA object with bootstrap resampling by
specifying the `B` argument, this is needed for statistical testing:

```{r boot}
B <- 25
numerator <- edma_fit(x1[1:25,,], B=B)
denominator <- edma_fit(x2[1:25,,], B=B)
```

The form difference matrix is defined as the ratio of the
two Euclidean distances form matrices based on the mean forms from the
numerator and denominator model objects:

```{r formdiff}
fd <- formdiff(numerator, denominator)
str(fd)
```

The global $T$-test assesses if the form difference is significant:

```{r ttest}
et <- edma_test(numerator, denominator)
et
plot(et)
```

The local test is achieved by looking at the stacked form difference,
where the bootstrap based `lower` and `upper` confidence limits are
also provided with the column `inside` indicating if the
`dist`ance is inside these limits:

```{r fdm}
fdm <- edma_fdm(numerator, denominator, 
    sort=TRUE, level=0.95)
head(fdm)
tail(fdm)
plot(fdm)
```

## Comparing 2 sets of 2D landmark data

```{r functions,include=FALSE}
edma_plot2d_data <- function(x, dims=1:2, landmarks=NULL, ...) {
    d <- x$data
    n <- length(d)
    if (length(dims) != 2L)
      stop("dims length must be 2")
    for (i in seq_len(n)) {
        d[[i]] <- d[[i]][,dims]
        d[[i]] <- t(t(d[[i]]) - colMeans(d[[i]]))
    }
    X <- do.call(rbind, d)
    L <- landmark_names(x)
    if (is.null(landmarks))
        landmarks <- L
    if (is.numeric(landmarks))
        landmarks <- L[landmarks]
    K <- length(L)
    y <- list()
    for (j in seq_len(K)) {
        y[[L[j]]] <- t(sapply(d, function(z) z[j,]))
    }
    rx <- range(X[,1L])
    ry <- range(X[,2L])
    plot(0, type="n", xlim=rx, ylim=ry, 
        xlab=colnames(X)[1L], ylab=colnames(X)[2L], main=x$name)
    for (j in seq_len(K)) {
        if (L[j] %in% landmarks) {
            points(y[[j]], col=j)
            points(matrix(colMeans(y[[j]]), nrow=1), pch=3, col=j, cex=2)
        }
    }
    invisible(x)
}
edma_plot2d_npfit <- function(x, add=FALSE, ...) {
    M <- Meanform(x)
    SKS <- SigmaKstar(x)
    sig <- 2 * sqrt(diag(SigmaKstar(x)))
    cex <- 2.5 * (0.5 + sig) / max(0.5 + sig)
    if (!add) {
        plot(M, cex=cex, ...)
    } else {
        points(M, cex=cex, ...)
    }
    invisible(x)
}
```

Using data from Lele and Richtsmeier 1991, with 2D landmarks visualized:

```{r boot2d}
file_a <- system.file("extdata/apert/ALLAPT4.xyz", 
    package="EDMAinR")
(a <- read_xyz(file_a))

file_b <- system.file("extdata/apert/ALLNORM4.xyz", 
    package="EDMAinR")
(b <- read_xyz(file_b))

op <- par(mfrow=c(1,2))
edma_plot2d_data(a)
edma_plot2d_data(b)
par(op)
```

Here are the two EDMA fit objects and the mean form visualized in 2D.
Sizes of circles are proportional to diagonal values in $\hat{\Sigma}^{\ast}_K$:

```{r fit2d}
numerator <- edma_fit(a, B=B)
denominator <- edma_fit(b, B=B)
```

```{r plot2dfit,echo=FALSE}
Ma <- Meanform(numerator)
Mb <- Meanform(denominator)
edma_plot2d_npfit(numerator, col=2, 
    xlim=range(Ma[,1], Mb[,1]),
    ylim=range(Ma[,2], Mb[,2]))
edma_plot2d_npfit(denominator, col=4, add=TRUE)
for (i in 1:10){
    lines(c(Ma[i,1], Mb[i,1]), c(Ma[i,2], Mb[i,2]), col="grey")
}
legend("topright", bty="n", pch=21, col=c(2,4),
    legend=c("Apert", "Normal"))
```

We had to drop a normal child, and maybe as a result, the observed T statistic
is lower than the published result. The bootstrap distribution seems
similar, but now a lot more T's are higher than the observed T.

```{r et2d}
(et <- edma_test(numerator, denominator))
plot(et)
```

Form difference matrix visualized:

```{r fdm2d}
fdm <- edma_fdm(numerator, denominator, 
    sort=TRUE, level=0.95)
plot(fdm)
```

## Comparing relative growth

We will use the Cabus data sets from the 'Purple Book':

```{r gdata}
file_a1 <- system.file("extdata/purplebook/cebusage1.xyz", 
    package="EDMAinR")
a1 <- read_xyz(file_a1)
file_a2 <- system.file("extdata/purplebook/cebusage6.xyz", 
    package="EDMAinR")
a2 <- read_xyz(file_a2)
a1
a2
```

Estimate the nonparametric forms and then the growth matrix:

```{r gm}
fit_a1 <- edma_fit(a1, B=B)
fit_a2 <- edma_fit(a2, B=B)

gm <- edma_gm(fit_a1, fit_a2, sort=TRUE, level=0.95)
head(gm)
tail(gm)
```

The `edma_gdm` function works similarly to the `edma_fdm` function,
but instead of 2 fitted objects it takes 4 inputs: a1, a2, b1, and b2.


## References

Lele, S. R., and Richtsmeier, J. T., 1991.
Euclidean distance matrix analysis: A coordinate‐free approach for 
comparing biological shapes using landmark data.
American Journal of Physical Anthropology 86:415--27.
DOI: [10.1002/ajpa.1330860307](https://doi.org/10.1002/ajpa.1330860307).

Lele, S. R., and Richtsmeier, J. T., 1995.
Euclidean Distance Matrix Analysis: Confidence Intervals for
Form and Growth Differences.
American Journal of Physical Anthropology 98:73--86.

Hu, L., 2007. Euclidean Distance Matrix Analysis of Landmarks Data:
Estimation of Variance. Thesis, Master of Science in Statistics,
Department of Mathematical and Statistical Sciences, 
University of Alberta, Edmonton, Alberta, Canada. Pp. 49.
