---
title: "EDMA growth matrix"
author: "Peter Solymos"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EDMA growth matrix}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup,include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(429)
```

# Introduction

This tutorial explains how to calculate growth matrix based
on 2 EDMA data objects with homologous landmarks.
We will use data sets where landmarks were measured on specimens
of 2 different ages:

```{r}
library(EDMAinR)

file_a1 <- system.file("extdata/purplebook/cebusage1.xyz",
    package="EDMAinR")
a1 <- read_xyz(file_a1)
file_a2 <- system.file("extdata/purplebook/cebusage6.xyz",
    package="EDMAinR")
a2 <- read_xyz(file_a2)
a1
a2
```

# Estimating the growth matrices

We first estimate the mean forms (no bootstrap replicates are necessary).

```{r}
fit_a1 <- edma_fit(a1)
fit_a2 <- edma_fit(a2)
```

Growth matrices ($GM$) are formed as pairwise Euclidean distances
between landmarks from EDMA fit objects using the estimated mean forms
from objects `a1` and `a2`.

The growth matrix is calculated as the ratio of form matrices
($FM$) from the numerator and denominator objects: $FDM(a1,a2) = FM(a2)/FM(a1)$.


```{r}
gm <- edma_gm(a1=fit_a1, a2=fit_a2, B=10)
gm
```

## Global T-test

The global T-test is based on the pairwise distances in the $FM$s, 
taking the max/min ratio of the distances.
This is then done for all the `B` replicates, which provides the null
distribution. The $p$ value for the T-test is calculated
as the number of replicates where the bootstrap T value was higher than 
or equal to the observed T value, divided by `B`.

`T_test` provides a summary of the test, `plot_Ttest` helps to visualize
the null distribution (histogram) and the observed T statistic
(vertical line):

```{r fig.width=7,fig.height=5,out.width='100%'}
T_test(gm)
plot_Ttest(gm)
```

## Local testing

The local testing is done based on the
confidence intervals using the stacked $GM$s from the bootstrap.
Here the distribution charactersizes the $GM$ and not the $FM$.
The confidence level can be changed through the `level` argument.

```{r}
head(confint(gm))
head(get_gm(gm))
head(get_gm(gm, sort=TRUE, decreasing=TRUE))
head(get_gm(gm, sort=TRUE, decreasing=FALSE))
```

The `plot_ci` function shows the pairwise differences and confidence intervals:

```{r fig.width=7,fig.height=7,out.width='100%'}
plot_ci(gm)
```


# Influential landmarks

Influence is calculated similarly to $FDM$:

```{r}
get_influence(gm)
plot(get_influence(gm))
```

# Ordination and clustering for specimens


```{r fig.width=7,fig.height=7,out.width='100%'}
plot_ord(gm)
```

The dendrogram leaves (speciemn labels) are also colored by groups:

```{r fig.width=7,fig.height=7,out.width='100%'}
plot_clust(gm)
```


# Visualizing landmarks

The 2D and 3D plots produce a plot of the mean form
from the reference object ('prototype').
The color intensity for the landmarks (dots) is associated with the `Tdrop`
influence value (larger the difference, the more intensive the color;
red by default).
Lines between the landmarks represent distances.
We use the diverging palettes:
<1 differences are colored blue (1st half of the palette), >1 differences 
are colored red (2st half of the palette).

```{r fig.width=7,fig.height=5,out.width='100%'}
plot_2d(gm, cex=2)
```



```{r fig.width=7,fig.height=5,out.width='100%'}
library(rgl)
xyz <- plot_3d(gm)
text3d(xyz, texts=rownames(xyz), pos=1) # this adds names
decorate3d() # this adds the axes
rglwidget(width = 600, height = 600, reuse = FALSE)
```

# Growth difference matrix

Growth difference matrix ($GDM$) is calculated as 
$GDM(A1,A2,B1,B2) = GM(B1,B2) / GM(A1,A2)$.

To be added later: example data set needed.


